---
title: "ChIP_analysis"
author: "TC"
date: "2023-11-27"
output: html_document
---

```{r libraries}
library(dplyr)
library(ggplot2)
library(DiffBind)
library(profileplyr)
```

```{r directories}
input <- "D:/path/to/input"
output <- "D:/path/to/output/"
figures <- "D:/path/to/figures/"
```

```{r modify the sample per what you have}
samples <- data.frame(
    SampleID = c("JIA_Active1", "JIA_Active2", "JIA_Active_2DG1", "JIA_Active_2DG2"),
    #Tissue = c("Tissue1", "Tissue1", "Tissue2", "Tissue2"),
    #Factor = c("Factor1", "Factor1", "Factor1", "Factor1"),
    Condition = c("Control", "Control", "Treatment", "Treatment"),
    Treatment = c("Untreated", "Untreated", "2DG", "2DG"),            # THIS IS NEEDED FOR CONTRAST LATER
    Replicate = c(1, 2, 1, 2),
    bamReads = c("D:/Enric_1/bams/JIAActive1_dedup.bam", "D:/Enric_1/bams/JIAActive2_dedup.bam", 
                 "D:/Enric_1/bams/JIAActive2DG1_dedup.bam", "D:/Enric_1/bams/JIAActive2DG2_dedup.bam"),
    Peaks = c(paste0(input,"JIAActive1_peaks.narrowPeak"), paste0(input,"JIAActive2_peaks.narrowPeak"),
              paste0(input,"JIAActive2DG1_peaks.narrowPeak"), paste0(input,"JIAActive2DG2_peaks.narrowPeak"))
)
```

```{r diffbind process}
# create the dba object 
dba.hc <- dba(sampleSheet=samples, peakFormat = "narrowPeak", scoreCol=5)
# count reads 
dba.hc <- dba.count(dba.hc)
# normalize the data
dba.hc <- dba.normalize(dba.hc)
# plot PCA- I was trying stuff
#dba.plotPCA(dba.hc, label=DBA_REPLICATE , method = DBA_DESEQ2, score=DBA_SCORE_RPKM)
dba.plotPCA(dba.hc, label=DBA_REPLICATE, th=1)

# if you want to save PCA uncomment the following
#pdf('D:/Enric_1/figures/PCA_all.pdf', width=7, height=8)
#dba.plotPCA(dba.hc, label=DBA_REPLICATE, th=1)
#dev.off()

# Add information - FRiP - if samples have really low FRiP score remove
info <- dba.show(dba.hc)
libsizes <- cbind(LibReads=info$Reads, FRiP=info$FRiP, PeakReads=round(info$Reads * info$FRiP))
rownames(libsizes) <- info$ID
libsizes

# set the contrast
dba.hc <- dba.contrast(dba.hc, minMembers=2, reorderMeta=list(Treatment="2DG"))
# Perform the differential binding analysis
dba.hc <- dba.analyze(dba.hc)
# retrieve the differentially bound sites
dba.hc.db <- dba.report(dba.hc, th = 1)         #th = 1 will give you all the peaks (significant and not)
db_all = data.frame(dba.hc.db)
# volcano plot
dba.plotVolcano(dba.hc, bReturnSites = T)
```

```{r create the volcano using ggplot2}
# modify the list
db_all$color <- ifelse((db_all$Fold < -1 | db_all$Fold > 1) & db_all$FDR <= 0.05, "red",
                       ifelse((db_all$Fold < -1 | db_all$Fold > 1) & db_all$FDR > 0.05, "green",
                              ifelse((db_all$Fold >= -1 & db_all$Fold <= 1) & db_all$FDR > 0.05, "gray", "blue")))

# Create a basic scatter plot --> change the labels for the following plot 
ggplot(db_all, aes(x=Fold, y=-log10(FDR), color=color)) +
  scale_color_manual(values = c("red" = "red", "blue" = "blue", "green" = "green", "gray" = "gray")) +
  geom_point(size=0.8) +
  theme_minimal() +
  geom_vline(xintercept = c(-1,1), linetype="dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype="dashed", color = "black") +
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5)) +
  annotate("text", x=1.7, y=13, label=paste0("HC act (",sum(db_all$color=="red" & db_all$Fold > 1),")"), color="black") +
  annotate("text", x=-2, y=13, label=paste0("HC 2DG (",sum(db_all$color=="red" & db_all$Fold < -1),")"), color="black") +
  annotate("text", x=0, y=13, label= sum(db_all$color=="blue"), color="black") +
  annotate("text", x=0, y=0.75, label= sum(db_all$color=="gray"), color="black") +
  annotate("text", x=-2, y=0.75, label= sum(db_all$color=="green" & db_all$Fold < -1), color="black") +
  annotate("text", x=1.7, y=0.75, label= sum(db_all$color=="green" & db_all$Fold > 1), color="black") +
  labs(title = 'HC active vs HC 2DG peaks,  FDR <= 0.05', x = 'log2 FC', y = '-log10(FDR)')

# save if you want
#ggsave(paste0(figures,"HC_act_HC2DG_volcano_with_values.pdf"),last_plot(), height = 6, width=8)

```

```{r Moving on to ChIPSeeker}
# libraries
library(ChIPseeker)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)      # make sure this is the correct database
#library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)

# create a dataframe
df <- as.data.frame(dba.hc.db)
# add the chr infront of the number (e.g. "7" --> "chr7")
df$seqnames <- paste0("chr",df$seqnames)

# separate the peaks based on FDR (similar to the heatmap previously)
gained_peaks <- df[df$FDR <= 0.05 & df$Fold >= 1, ]
lost_peaks <- df[df$FDR <= 0.05 & df$Fold <= -1, ]
common_peaks <- df[df$Fold > -1 & df$Fold < 1, ]
not_significant <- df[df$FDR > 0.05 & df$Fold > -1 & df$Fold < 1,]

# plot the numbers of the different dfs - sanity check 
nrow(gained_peaks)
nrow(lost_peaks)
nrow(common_peaks)
nrow(not_significant)

# we don't really need the rest of the columns so keep only those 
gained_peaks <- gained_peaks[c(1:3)]
lost_peaks <- lost_peaks[c(1:3)]
not_sig <- not_significant[c(1:3)]

# make all of those GenomicRanges objects
gained_peaks_gr <- makeGRangesFromDataFrame(gained_peaks)
lost_peaks_gr <- makeGRangesFromDataFrame(lost_peaks)
not_sig_gr <- makeGRangesFromDataFrame(not_significant)

# creating annotations. This will match each peak to a corresponding element close to it
# e.g. promoter, intron, 5 UTR etc. Changing tssRegions affects only promoter. You will now
# have more information regarding the relative position of the promoter. 
peak.anno_gained = annotatePeak(
  peak = gained_peaks_gr,
  tssRegion = c(-3000, 3000),
  TxDb = TxDb.Hsapiens.UCSC.hg19.knownGene,
  annoDb = "org.Hs.eg.db"
)

# same for lost regions
peak.anno_lost = annotatePeak(
  peak = lost_peaks_gr,
  tssRegion = c(-3000, 3000),
  TxDb = TxDb.Hsapiens.UCSC.hg19.knownGene,
  annoDb = "org.Hs.eg.db"
)

# save the GR object as dataframe if you want
#dfPA_gained= as.data.frame(peak.anno_gained)
#dfPA_lost = as.data.frame(peak.anno_lost)

# export the files 
#write.table(dfPA_gained, file = paste0(output,"gained_sites.csv"), sep=",", col.names=TRUE, row.names=FALSE, quote=FALSE )
#write.table(dfPA_lost, file = paste0(output,"lost_sites.csv"), sep=",", col.names=TRUE, row.names=FALSE, quote=FALSE )
```

```{r plotting using ChIPSeeker}
# make the annotation list for plotting
peakAnnoList <- c(peak.anno_gained, peak.anno_lost)
# provide names for the lists
names(peakAnnoList) <- basename(c("gained","lost"))

###### CHANGE LABEL NAMES ########
# plot 1 feature representation in png
png(filename=paste0(figures,"FR_HC_HC2DG.png"), units="in", width=7, height=4, res=300)
plotAnnoBar(peakAnnoList, title = "Feature Representation HC vs HC 2DG")
dev.off()
# and in pdf
pdf(file=paste0(figures,"FR_HC_HC2DG.pdf"), width=7, height=4)
plotAnnoBar(peakAnnoList, title = "Feature Representation HC vs HC 2DG")
dev.off()

# plot 2 distance to TSS in png
png(filename=paste0(figures,"TSS_HC_HC2DG.png"), units="in", width=7, height=4, res=300)
plotDistToTSS(peakAnnoList, title = "Distribution of transcription factor-binding loci relative to TSS")
dev.off()
# and in pdf
pdf(file=paste0(figures,"TSS_HC_HC2DG.pdf"), width=7, height=4)
plotDistToTSS(peakAnnoList, title = "Distribution of transcription factor-binding loci relative to TSS")
dev.off()
```

I stop at this point where we get the feature representation and distance from TSS plots. However we can continue and run GO/KEGG/etc enrichment analysis. I will provide the script I have but keep in mind it is not curated because I cannot find the required files in my computer.

```{r}
# Get the entrez IDs
entrez <- nanog_annot$geneId

# Return the gene symbol for the set of Entrez IDs
annotations_edb <- AnnotationDbi::select(EnsDb.Hsapiens.v75,
                                         keys = entrez,
                                         columns = c("GENENAME"),
                                         keytype = "ENTREZID")

# Change IDs to character type to merge
annotations_edb$ENTREZID <- as.character(annotations_edb$ENTREZID)

# Write to file
nanog_annot %>% 
  left_join(annotations_edb, by=c("geneId"="ENTREZID")) %>% 
  write.table(file="results/Nanog_peak_annotation.txt", sep="\t", quote=F, row.names=F)

library(clusterProfiler)
# Run GO enrichment analysis 
ego <- enrichGO(gene = dfPA_JIA$ENSEMBL, 
                    keyType = "ENSEMBL", 
                    OrgDb = org.Hs.eg.db, 
                    ont = "BP", 
                    pAdjustMethod = "BH", 
                    qvalueCutoff = 0.05,
                    readable = TRUE)

png(filename=paste0(output,"GO_JIA.png"), units="in", width=7, height=8, res=300)
# Dotplot visualization
dotplot(ego, showCategory=20,  title = "GO JIA peaks",)
dev.off()

pdf(file=paste0(output,"GO_JIA.pdf"), width=7, height=8)
# Dotplot visualization
dotplot(ego, showCategory=20,  title = "GO JIA peaks",)
dev.off()

ekegg <- enrichKEGG(gene = dfPA_JIA$geneId, 
                    organism = "hsa",
                    pAdjustMethod = "BH", 
                    qvalueCutoff = 0.05)

png(filename=paste0(output,"KEGG_JIA.png"), units="in", width=7, height=8, res=300)
# Dotplot visualization
dotplot(ekegg, showCategory=20,  title = "KEGG JIA peaks",)
dev.off()

pdf(file=paste0(output,"KEGG_JIA.pdf"), width=7, height=8)
# Dotplot visualization
dotplot(ekegg, showCategory=20,  title = "KEGG JIA peaks",)
dev.off()


# Create a list with genes from each sample
genes = lapply(peakAnnoList, function(i) as.data.frame(i)$geneId)

# Run KEGG analysis
compKEGG <- compareCluster(geneCluster = genes, 
                         fun = "enrichKEGG",
                         organism = "human",
                         pvalueCutoff  = 0.05, 
                         pAdjustMethod = "BH")

png(filename=paste0(output,"comp_KEGG_Enrichr_HC_JIA.png"), units="in", width=6, height=12, res=300)
dotplot(compKEGG, showCategory = 20, title = "KEGG Pathway Enrichment Analysis")
dev.off()

pdf(file=paste0(output,"comp_KEGG_Enrichr_HC_JIA.pdf"), width=6, height=12)
dotplot(compKEGG, showCategory = 20, title = "KEGG Pathway Enrichment Analysis")
dev.off()
```

And we can also plot the tornado plots - also not curated and maybe you need to load another package
# check more for tornados
https://content.cruk.cam.ac.uk/bioinformatics/software/DiffBind/plotProfileDemo.html

```{r}
# Specify the file name, width, and height
png("tornado_hc_vs_jia_active_E.png", width = 4, height = 8, units = "in", res = 300)

profiles <- dba.plotProfile(dba.hc)
dba.plotProfile(profiles)

# Close the PNG device
dev.off()
```
